---
title: "data_processing_notes.Rmd"
author: "Perla Lopez Flores"
date: "2022-07-14"
output: html_document
---
## Introduction to dplyr

dplyr is an R package in the tidyverse. We can load the package using

```{r}

# load required libraries
install.packages("dplyr")
library(dplyr)
library(ggplot2)
?dplyr

# preview dataset
head(starwars)

```

### Filtering data 
We can filter data using  'filter()'. This allows us to subset observations (rows) based on their values (in columns).

Tips:

  * Be sure you spell the column name correctly (and the value name if it's a categorical variable). Remember, R is case-sensitive.
  * Be sure to use '==' when comparing observation. (Remember, '=' is an assignment operator).
  * You can use >, <, >= to compare numeric or categorical values (nominal variables are ranked alphabetically, while ordinal variables have a built-in rank).

```{r}

# or is |
filter(starwars, hair_color == "blond" | eye_color == "blue")
filter(starwars, hair_color %in% c("blond", "blonde"))

importantPlanets <- c("Tatooine", "Naboo", "Mustafar")
filter(starwars, homeworld %in% importantPlanets)

filter(starwars, hair_color == "blond" & species == "Human")
filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")

filter(starwars, mass >= 75, mass <= 100, hair_color == "brown", height > 170)

# ! is for "not"
filter(starwars, mass != 75, name < "Mace" )

filteredData <- filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")

# ranked data
head(diamonds) # displays first 6 rows
class(diamonds$cut) # gives you the specific type of data
summary(diamonds$cut) # gives you a count of each category or quartiles & summary statistics if numeric
summary(diamonds$carat)
head(diamonds$cut) # displays first 6 values (and levels)

filter(diamonds, cut > "Good")


## Ordering categorical data
unique(starwars$eye_color)
factor(starwars$eye_color, c("red", "orange", "gold", "green,yellow" , "blue", "black", ordered = T))

## Find all characters that are shorter than 100 cm
filter(starwars, height < 100, species != "Droid")

## Which characters were born between the years 100 and 200 (inclusive)
filter(starwars, birth_year >= 100, birth_year <= 200)
filter(starwars, between(birth_year, 100, 200)) # This is a faster way of looking for data in between numbers

## Which characters weight over 100kg but are shorter than 185cm
filter(starwars, mass > 100, height < 185)

## Which characters are missing a hair color?
filter(starwars, is.na(hair_color))

### Missing values
NA > 5 # It doesn't know
NA + 10
NA/2

mary_age <- NA
paul_age <- NA
mary_age == paul_age
mass <- NA
mass == 10 # returns NA
is.na(NA) # returns trie

```
### Arranging data
'arrange()' reorders rows. It does not remove any rows.
NA values are always at the end when you order bya column.
```{r}

# lowest to highest birth_year
arrange(starwars, birth_year)

# highest to lowest birth_year
arrange(starwars, desc(birth_year))

# categorical in alphabetical
arrange(starwars, hair_color)

# multiple columns
arrange(starwars, hair_color, birth_year)

### Practice!

## Arrange starwars characters to find the tallest characters and the shortest characters
arrange(starwars, height)
arrange(starwars, desc(height))

## Alphabetize the star wars characters by name
arrange(starwars, name)

## How could you use arrange() to sort all missing values to the start?
arrange(starwars, desc(is.na(height)), desc(is.na(mass)))

arrange(starwars, desc(is.na(starwars)))

```

### We can select certain columns in the dataset

'select()' allows us to retain only certain variables (columns).
It doesn't change the order, but it removes columns not named.

```{r}

select(starwars, hair_color, skin_color, eye_color)

head(starwars)
select(starwars, hair_color:eye_color) #returns every column between first:last

select(starwars, -hair_color)
select(starwars, -(hair_color:eye_color))

starwars_noColor <- select(starwars, -(hair_color:eye_color))
ggplot(starwars_noColor, aes(x = hair_color)) # error because we removed it

select(starwars, contains("color"))
select(starwars, ends_with("color"))
select(starwars, contains("_"))
select(starwars, starts_with("s"), ends_with("color"))

?select

rename(starwars, birthYear = birth_year) # can use rename to rename some of the columns, new -> old

starwars2 <- rename(starwars, birthYear = birth_year)
starwars2

### Practice!!

## Select out the homeworld and species for the starwars datatset. What question might this subset of variables help answer?
select(starwars, homeworld, species)

## Brainstorm as many ways as possible to select hair_color, eye_color, skin_color, and birth_year
select(starwars, ends_with("color"), ends_with("year"))

```

### Adding new columns
'mutate()' adds new columns to the end of your dataset.

```{r}

starwars_small <- select(starwars, height, mass, birth_year)
head(starwars_small)

starwars_small <- mutate(starwars_small, height_m = height/100)
mutate(starwars_small, bmi = mass/(height_m^2))

starwars_small <- select(starwars, height, mass, birth_year)
mutate(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

# to only keep new columns, use transmute
transmute(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

## using aggregate functions
prop_mass <- mutate(starwars_small, proportional_mass = mass/sum(mass, na.rm = T))
arrange(prop_mass, desc(proportional_mass))

```

### Summarizing and grouping data

'summarize()' collapse an entire column of data to a single value

```{r}

mutate(starwars, mean_mass = mean(mass, na.rm = T))
summarize(starwars, mean_mass = mean(mass, na.rm = T), mean_height = mean(height, na.rm = T))

mean(starwars$mass, na.rm = T)

species_masses <- summarize(group_by(starwars, species), mean_mass = mean(mass, na.rm = T), count = n()) # n returns the number of values in each species

arrange(species_masses, desc(mean_mass))

# the pipe operator %>%
# function(x, y) is the same as x %>% function(y)
# when using dplyr functions, generally always start with dataset
species_masses <- starwars %>% 
  group_by(species) %>%
  summarize(mean_mass = mean(mass, na.rm = T), count = n()) %>%
  arrange(desc(mean_mass))

species_masses

```

### Sampling a designated numbeer of rows
'sample_n()' allos us to sample a random number of rows from our dataset. Sampling is randomm.

```{r}

# 10 random rows
starwars_10rows <- sample_n(starwars,10)
starwars_10row

# 10% of rows selected randomly
nrow(starwars) # total number of rows is 87
starwars_10percent <- sample_frac(starwars, 0.1)
starwars_10percent # 9 rows is 10%

starwars %>%
  group_by(gender) %>%
  sample_n(4)

```

We can also take a "slice" of our dataset using 'slice()' and its related set of functions. It is not random.

```{r}

slice(starwars, 50:60) # Slice by index value (row number)

slice_head(starwars, n = 10) # first 10 rows by index
slice_tail(starwars, n = 10) # last 10 rows by index

slice_min(starwars, order_by = mass, n = 10) # rows with lowest 10 masses
slice_max(starwars, order_by = mass, n = 10) # rows with highest 10 masses

```

